\chapter*{감사의 글}
\pdfbookmark{Acknowledgments}{acknowledgments}

%Thanks to the countless authors and content producers who influenced my thinking
%on Bitcoin and the topics it touches. There are too many to list them all, but
%I’ll do my best to name a few.
비트코인에 대해 제 생각에 영향을 준 수 많은 저자와 콘텐츠 제작자에게 감사드립니다.
모두 나열하기에는 너무 많지만, 몇 분만 추려서 언급하겠습니다.


\begin{itemize}
	\begin{comment}
		\item Thanks to Arjun Balaji for the tweet which motivated me to write this.
		\item Thanks to Marty Bent for providing endless food for thought and entertainment. If you are not subscribed to Marty’s Bent and Tales From The Crypt, you are missing out. Cheers Matt and Marty for guiding us through the rabbit hole.
		\item Thanks to Michael Goldstein and Pierre Rochard for curating and providing the greatest Bitcoin literature via the Nakamoto Institute. And thank you for creating the Noded Podcast which influenced my philosophical views on Bitcoin substantially.
		\item Thanks to Saifedean Ammous for his convictions, savage tweets, and writing The Bitcoin Standard
		\item Thanks to Francis Pouliot for sharing his excitement about finding out about the timechain.
		\item Thanks to Andreas M. Antonopoulos for all the educational material he has put out over the years.
		\item Thanks to Peter McCormack for his honest tweets and the What Bitcoin Did podcast, which keeps providing great insights from many areas of the space.
		\item Thanks to Jannik, Brandon, Matt, Camilo, Daniel, Michael, and Raphael for providing feedback to early drafts of some lessons. Special thanks to Jannik who proofread multiple drafts multiple times.
		\item Thanks to Dhruv Bansal and Matt Odell for taking the time to discuss some of these ideas with me.
		\item Thanks to Guy Swann for producing an audio version of 21lessons.com.
		\item Thanks to Friar Hass for his spiritual support and guidance, and for taking the time to write a foreword for this book.
		\item Thanks to my wife for putting up with me and my obsessive nature.
		\item Thanks to my family for supporting me during both the good times and the bad.
		\item Last but not least, thanks to all the bitcoin maximalists, shitcoin minimalists, shills, bots, and shitposters which reside in the beautiful garden that is Bitcoin twitter.
	\end{comment}
	\item 이 글을 쓰도록 동기를 제공한 아준 발라지(Arjun Balaji)에게 감사드립니다.
	\item 생각과 여가를 위하여 끝없이 음식을 제공한 마티 벤트 (Marty Bent)에게 감사드립니다. 마티 벤트 (Marty’s Bent)와 크립토 이야기(Tales From The Crypt)를 구독해 주세요. 토끼굴을 통해 우리에게 안내해 준 맷과 마티에게 환호를 보냅니다.
	\item 나카모토 재단을 통해 최고의 비트코인 문헌을 제공한 마이클 골드스타인(Michael Goldstein)과 피에르 로차드(Pierre Rochard)에게 감사드립니다. 그리고 비트코인에 대한 철학적 견해에 상당한 영향을 준 노디드(Noded) 팟캐스트를 만들어주셔서 감사합니다.
	\item 그의 신념, 잔인한 트윗과 비트코인 스탠다드를 집필해 준 사이페딘 아모스(Saifedean Ammous)에게 감사드립니다.
	\item 타임체인에 대해 알게 된 기쁨을 공유해 준 프랑시스 폴리엇(Francis Pouliot)에게 감사드립니다.
	\item 수년에 걸쳐 모든 교육자료를 제공한 안드레아스 안토노풀로스(Andreas M. Antonopoulos)에게 감사드립니다.
	\item 솔직한 트윗과 공간의 여러 영역에서 훌륭한 통찰력을 지속적으로 제공한 What Bitcoin Did 팟캐스트의 피터 맥코맥(Peter McCormack)에게 감사드립니다.
	\item 일부 교훈 초기에 피드백을 준 재닉(Jannik), 브랜든(Brandon), 맷(Matt), 카밀로(Camilo), 다니엘(Daniel), 마이클(Michael), 라파엘(Raphael)에게 감사드립니다. 어려 초안을 여러 번 교정해 준 재닉에게 특별히 감사드립니다.
	\item 시간을 내어 아이디어에 대해 논의해 준 데룹 반살(Dhruv Bansal)과 맷 오델(Matt Odell)에게 감사드립니다.
	\item 21lessons.com의 오디오 버전을 제작해 준 가이 스완(Guy Swann)에게 감사드립니다.
	\item 영적 지원과 인도를 배풀고 시간을 내어 이 책의 서문을 써 준 하스(Hass) 신부에게 감사드립니다.
	\item 나의 강박적인 성격을 참아준 아내에게 감사드립니다.
	\item 좋을 때나 나쁠 때나 저를 지지해 준 가족들에게 감사드립니다.
	\item 마지막으로 비트코인 트위터라는 아름다운 정원에 상주하는 모든 비트코인 맥시멀리스트, 똥코인 미니멀리스트, 사기꾼들, 봇, 똥포스터에게 감사드립니다.
\end{itemize}


%And finally, thank you for reading this. I hope you enjoyed it as much as I did enjoy writing it.
그리고 마지막으로 이 글을 읽어주셔서 감사합니다. 
제가 글을 쓰면서 즐거웠던 만큼 여러분도 즐거우셨기를 바랍니다.
